Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-06-21 abstract.tex

\begin{abstract}
Across almost all scientific disciplines, instruments that record experimental data and methods required for storage and data analysis are rapidly increasing in complexity.
As a result of these rapid developments, scientific communities need to adapt on shorter time scales than traditional university curricula can evolve. New modes of knowledge transfer are therefore needed.
On the other hand, the broad applicability of data science tools to a range of problems has generated new opportunities to foster exchange of ideas and computational workflows across disciplines.
In recent years, hack weeks have emerged as a novel and effective tool for fostering these exchanges and providing training in modern data analysis workflows.
While there are variations in hack week implementation, all events consist of a common core of three components: tutorials in state-of-the-art methodology, peer-learning and project work in a collaborative environment.
In this paper, we present the concept of a hack week in the larger context of scientific meetings.
We motivate the need for such an event and present its strengths and challenges.
We find that hack weeks are successful at cultivating collaboration and facilitating the exchange of knowledge.
Participants self-report that these events help them both in their day-to-day research as well as their careers.
Based on our results, we conclude that hack weeks present an effective, easy-to-implement, fairly low-cost tool to positively impact data analysis literacy in academic disciplines, foster collaboration and cultivate best practices.
\end{abstract}

>>>>

\begin{abstract}
Across almost all scientific disciplines, instruments and methods for recording, storing and analyzing data are rapidly increasing in complexity.
Numerous data science tools are in development that aim to manage this complexity through development of open, reproducible workflows and the exchange of ideas across disciplines.
Successful navigation of this new data science landscape requires training in new programming languages and frameworks, as well as immersion in new modes of interaction that foster data sharing and collaborative software development. 
Unfortunately, learning these skills from traditional university curricula can be challenging because most courses are not designed to evolve on time scales that can keep pace with rapidly shifting data science methods.
Here we present the concept of a hack week as a novel and effective model offering opportunitites for networking and community building, education in state-of-the-art data science methods and immersion in collaborative project work.
We find that hack weeks are successful at cultivating collaboration and facilitating the exchange of knowledge.
Participants self-report that these events help them both in their day-to-day research as well as their careers.
Based on our results, we conclude that hack weeks present an effective, easy-to-implement, fairly low-cost tool to positively impact data analysis literacy in academic disciplines, foster collaboration and cultivate best practices.
\end{abstract}

### 2018-06-21 results.tex

One important question is whether participants from underrepresented groups thrive at hack weeks, or whether their full participation is impeded. Significant differences between minorities and non-minorities, even on a self-reported scale, for questions related to learning outcomes, teaching or network building, would indicate that improvements in workshop facilitation and structure may be required to allow a wide range of participants to participate fully.
Based on our surveys, we find no significant dependence of self-reported learning outcomes on gender identity or race/ethnicity ($p > 0.0007$). Similarly, minority participants with respect to gender identity and race or ethnicity do not significantly differ in their answers with respect to teaching outcomes for any of the hack weeks ($p > 0.0007$). Attitudes towards building valuable connections at all three hack weeks or the value of their contributions to their hack teams also do not differ with respect to gender or racial/ethnic identity. For GHW, there is a weak indication that participants from racial/ethnic minorities respond more positively when asked about building connections ($p=0.04$) with a medium to large effect size ($\phi_c = 0.4$; $\mathrm{dof}_{\phi_c} = 0.97$), while for AHW, responses regarding the value of contributions to hack teams may be spread more widely for participants from racial/ethnic minorities than for Caucasian participants ($p = 0.01$; $\phi_c = 0.4$, $\mathrm{dof}_{\phi_c} = 0.98$). Equivalence tests for the hypothesis of a small or non-existent effect reveal systematically small, though non-significant, $p$-values in the range of $p=0.02-0.05$ for AHW for all four questions above in conjunction with gender or ethnic/racial identity, while for GHW and NHW $p > 0.05$ for the same questions (note that the sample size for GHW and NHW was smaller by a factor of 2 compared to AHW). These results, while not a decisive exclusion of an effect of race/ethnicity or gender on hack week participation, provide an initial indication that our careful facilitation strategies may be effective in fostering participation, and that future work on how demographics interact with hack week attendance may be fruitful.

We find that the hack weeks have been largely successful in promoting positive attitudes towards reproducibility and open science: at all three events, the majority reports that the hack week has made them more comfortable with open science  (GHW: 97\%; NHW: 95\%; AHW: 72\%, Figure \ref{fig:survey}, (h)), and more than 85\% of all participants (AHW: 86\%, GHW: 94\%, NHW: 95\%; Figure \ref{fig:survey}, (g)) put code or data created at the hack week into a public repository.
%The overall behaviour hints at how general conventions and attitudes differ in different fields.
%Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science (Figure \ref{fig:survey}, (j)): again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, followed by neuroscience (95\%) and astronomy (72\%).
%Overall, our results indicate that hack weeks are effective at addressing persisting doubts about making research open and reproducible.
While the focus on open science is not necessarily a required component of a hack week, it aligns naturally with many of the goals and values commonly promoted at hack weeks, such as production of open-source software and data sharing. In some fields, especially where ethical issues around data sharing and privacy are relevant, this might not be a desired focus of the hack week, or might be replaced or augmented by a discussion of ethical considerations.

>>>>

One important question is whether participants from underrepresented groups thrive at hack weeks, or whether their full participation is impeded. Significant differences between minorities and non-minorities, even on a self-reported scale, for questions related to learning outcomes, teaching or network building, would indicate that improvements in workshop facilitation and structure may be required to allow a wide range of participants to participate fully.

Based on our surveys, we find no significant dependence of self-reported learning outcomes on gender identity or race/ethnicity ($p > 0.0007$). Similarly, minority participants with respect to gender identity and race or ethnicity do not significantly differ in their answers with respect to teaching outcomes for any of the hack weeks ($p > 0.0007$). Attitudes towards building valuable connections at all three hack weeks or the value of their contributions to their hack teams also do not differ with respect to gender or racial/ethnic identity. For GHW, there is a weak indication that participants from racial/ethnic minorities respond more positively when asked about building connections ($p=0.04$) with a medium to large effect size ($\phi_c = 0.4$; $\mathrm{dof}_{\phi_c} = 0.97$), while for AHW, responses regarding the value of contributions to hack teams may be spread more widely for participants from racial/ethnic minorities than for Caucasian participants ($p = 0.01$; $\phi_c = 0.4$, $\mathrm{dof}_{\phi_c} = 0.98$). Equivalence tests for the hypothesis of a small or non-existent effect reveal systematically small, though non-significant, $p$-values in the range of $p=0.02-0.05$ for AHW for all four questions above in conjunction with gender or ethnic/racial identity, while for GHW and NHW $p > 0.05$ for the same questions (note that the sample size for GHW and NHW was smaller by a factor of 2 compared to AHW). These results, while not a decisive exclusion of an effect of race/ethnicity or gender on hack week participation, provide an initial indication that our careful facilitation strategies may be effective in fostering participation, and that future work on how demographics interact with hack week attendance may be fruitful.

We find that the hack weeks have been largely successful in promoting positive attitudes towards reproducibility and open science: at all three events, the majority reported that the hack week has made them more comfortable with open science  (GHW: 97\%; NHW: 95\%; AHW: 72\%, Figure \ref{fig:survey}, (h)), and more than 85\% of all participants (AHW: 86\%, GHW: 94\%, NHW: 95\%; Figure \ref{fig:survey}, (g)) put code or data created at the hack week into a public repository.
%The overall behaviour hints at how general conventions and attitudes differ in different fields.
%Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science (Figure \ref{fig:survey}, (j)): again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, followed by neuroscience (95\%) and astronomy (72\%).
%Overall, our results indicate that hack weeks are effective at addressing persisting doubts about making research open and reproducible.
While the focus on open science is not necessarily a required component of a hack week, it aligns naturally with many of the goals and values commonly promoted at hack weeks, such as production of open-source software and data sharing. In some fields, especially where ethical issues around data sharing and privacy are relevant, this might not be a desired focus of the hack week, or might be replaced or augmented by a discussion of ethical considerations.

### 2018-06-21 supplementary_material.tex

Participants were asked to respond to statements regarding these topics using a six-point Likert-type scale. All questions were anonymously recorded. No responses were discarded, and no pre-processing was performed on the data. We tested for correlations between demographic characteristics (independent variables), questioned responses (dependent variables) using a standard $\chi^2$ test and computed the effect sizes via a bias-corrected version of Cram\'{e}r's V \citep{cramer1946,bergsma2013}, denoted $\phi_c$ in the manuscript. For a fairly permissive significant threshold of $\alpha < 0.05$, we require a $p$-value of $p < 0.0007$, corrected for $N_\mathrm{trials} = 36$ trials, to claim a significant effect as well as at least a small effect size according to~\citep{cohen1988}. The full analysis procedure is available online\footnote{See the repository: \url{https://github.com/uwescience/HackWeek-Writeup}}.

@@

\begin{tabularx}{18cm}{p{3.6cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}p{0.42cm}}

@@

\textbf{Question} & \multicolumn{6}{c}{\textbf{Astro Hack Week}} & \multicolumn{6}{c}{\textbf{GeoHackWeek}} & \multicolumn{6}{c}{\textbf{NeuroHackWeek}} \\ \midrule
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ \\ \midrule

>>>>

Participants were asked to respond to statements regarding these topics using a six-point Likert-type scale. All questions were anonymously recorded. No responses were discarded, and no pre-processing was performed on the data. We test for correlations between demographic characteristics (independent variables) and question responses (dependent variables) using a standard $\chi^2$ test and compute the effect sizes via a bias-corrected version of Cram\'{e}r's V \citep{cramer1946,bergsma2013}, denoted $\phi_c$ in the manuscript. For a fairly permissive significant threshold of $\alpha < 0.05$, we require a $p$-value of $p < 0.0007$, corrected for $N_\mathrm{trials} = 36$ trials, to claim a significant effect as well as at least a small effect size according to~\citep{cohen1988}.

We note that while a non-significant $p$-value can reject a hypothesis, this is not automatically evidence for the corresponding null hypothesis, since a lack of statistical power could also originate in a too small sample size. For cases where the null hypothesis corresponds to the scientific hypothesis of interest, we perform an equivalence test based on the Cram\'{e}r's V effect size following~\citep{}. We generate 10,000 random samples via bootstrap for each response variable and compute the corresponding effect size in order to generate 90\% confidence intervals around the measured Cram\'{e}r's V. For each hypothesis, we determine an equivalence bound for Cram\'{e}r's V for a medium-strength effect size based on \citet{cohen1988} given the number of degrees of freedom and calculate the $p$-value that the measured effect size could extend above\footnote{Note that because Cram\'{e}r's V is defined on the interval $[0,1]$, we only require an upper bound} this equivalence bound. If the latter hypothesis can be rejected, this indicates that the observed effect falls within the equivalence bounds and is statistically smaller than any effect deemed worthwhile.

The full analysis procedure is available online\footnote{See the repository: \url{https://github.com/uwescience/HackWeek-Writeup}}.

@@

\begin{tabularx}{24cm}{p{3.2cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}p{0.4cm}}

@@

\textbf{Question} & \multicolumn{9}{c}{\textbf{Astro Hack Week}} & \multicolumn{9}{c}{\textbf{GeoHackWeek}} & \multicolumn{9}{c}{\textbf{NeuroHackWeek}} \\ \midrule
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $\mathrm{CI}_{\mathrm{L}}$ & $\mathrm{CI}_{\mathrm{U}}$ & $p_{\mathrm{eq}}$ 
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ &$\mathrm{CI}_{\mathrm{L}}$ & $\mathrm{CI}_{\mathrm{U}}$ & $p_{\mathrm{eq}}$ 
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $\mathrm{CI}_{\mathrm{L}}$ & $\mathrm{CI}_{\mathrm{U}}$ &$p_{\mathrm{eq}}$  \\ \midrule
